Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2020-09-03 journal paper/content/sec-intro.tex
More small edits.

The main high-level question is: 
%that we focus on in this paper is:
%\begin{align}\label{eq:main-Q}
\textbf{How does competition incentivize the adoption of better exploration algorithms?}
%\end{align}
This translates into a number of more concrete questions. While it is commonly assumed that better technology always helps, is this so under competition? Does increased competition lead to higher consumer welfare? How significant are ``data feedback loops" --- when more data leads to more users, which leads to even more data, etc. --- and how they relate to the anti-trust considerations? 
%Finding formalizations that admit meaningful answers is a major part of the overall challenge.

We offer a mix of theoretical results and numerical simulations, in which we study complex interactions between platforms' learning dynamics and users' self-interested behavior. The choice of a particular technology (exploration algorithm) is no longer an abstract, static choice with a predetermined outcome for the platform. Instead, we model the algorithms explicitly, and investigate how they play out in competition over an extended period of time.

@@

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a bandit algorithm; as such, it adjusts itself to the observed rewards, but not to the competition. This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning. Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical. We draw on the machine learning literature to compare bandit algorithms to one another in the absence of competition, so that we can talk about \emph{better} bandit algorithms in a principled way.%
\footnote{Comparisons between bandit algorithms are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. Generally,  ``better" algorithms are better in the long run, but could be worse initially. More on this in Preliminaries.} One baseline is \emph{greedy algorithms} that do not explicitly explore; in the absence of competition, it is known to be terrible for a wide variety of problem instances.

@@

%\begin{itemize}
\textbf{(i)} We compare \HardMax and \HardMaxRandom decision rules. We find that the greedy algorithm often wins under \HardMax, with a strong evidence of the ``death spiral" effect mentioned earlier.
 As predicted by the theory, better algorithms prevail if the expected number of ``random" users is sufficiently large. However, this effect is negligible for smaller parameter values.

@@

\xhdr{Challenges.}
Much of the challenge in the theoretical results, both conceptual and technical, was in setting up the model and the theorems. Apart from zeroing in on the \TheoryModel, it was crucial to interpret the results and intuitions from the literature on multi-armed bandits so as to formulate meaningful and productive assumptions on bandit algorithms and Bayesian priors.

The numerical investigation is quite challenging even with a stylized model such as ours. An ``atomic experiment" is a competition game between a given pair of bandit algorithms, in a given competition model, on a given instance of a multi-armed bandit problem (and each such experiment is run many times to reduce variance).
Accordingly, we have a three-dimensional space of atomic experiments one needs to run and interpret: \{pairs of algorithms\} x \{competition models\} x \{bandit instances\}, and we are looking for findings that are consistent across this entire space. It is essential to keep each of the three dimensions small yet representative. In particular, we need to capture a huge variety of bandit instances with only a few representative examples. Further, we need a succinct and informative summarization of results within one atomic experiment and across multiple experiments (\eg see Table~\ref{sim_table}).

>>>>

The main high-level question is: 
%that we focus on in this paper is:
%\begin{align}\label{eq:main-Q}
\textbf{How does competition incentivize the adoption of better exploration algorithms?}
%\end{align}
This translates into a number of more concrete questions. While it is commonly assumed that better technology always helps, is this so under competition? Does increased competition lead to higher consumer welfare? How significant are ``data feedback loops" --- when more data leads to more users, which leads to even more data, etc. --- and how they relate to the anti-trust considerations? 
%Finding formalizations that admit meaningful answers is a major part of the overall challenge.
We offer a mix of theoretical results and numerical simulations, in which we study complex interactions between platforms' learning dynamics and users' self-interested behavior.

%The choice of a particular technology (exploration algorithm) not an abstract, static choice with a predetermined outcome for the platform. Instead, we model the algorithms explicitly, and investigate how they play out in competition over an extended period of time.

@@

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a bandit algorithm; as such, it adjusts itself to the observed rewards, but not to the competition. This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning. Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical. We draw on the machine learning literature to compare bandit algorithms to one another in the absence of competition, so that we can talk about \emph{better} bandit algorithms in a principled way.%
\footnote{Comparisons between bandit algorithms are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. Generally,  ``better" algorithms are better in the long run, but could be worse initially. More on this in Preliminaries.} One baseline is algorithms that do not explicitly explore (\emph{greedy algorithms}); in the absence of competition, they are known to be terrible for a wide variety of problem instances.

@@

%\begin{itemize}
\textbf{(i)} 
%We compare \HardMax and \HardMaxRandom decision rules. 
We find that the greedy algorithm often wins under the \HardMax decision rule, with a strong evidence of the ``death spiral" effect mentioned earlier.
 As predicted by the theory, better algorithms prevail under \HardMaxRandom if the expected number of ``random" users is sufficiently large. However, this effect is negligible for smaller parameter values.

@@

\xhdr{Challenges.}
Much of the difficulty in the theoretical results, both conceptual and technical, was in setting up the model and the theorems. Apart from zeroing in on the \TheoryModel, it was crucial to interpret the results and intuitions from the literature on multi-armed bandits so as to formulate meaningful and productive assumptions on bandit algorithms and Bayesian priors.

The numerical investigation is quite challenging even with a stylized model such as ours. An ``atomic experiment" is a competition game between a given pair of bandit algorithms, in a given competition model, on a given instance of a multi-armed bandit problem (and each such experiment is run many times to reduce variance).
Accordingly, we have a three-dimensional space of atomic experiments one needs to run and interpret: \{pairs of algorithms\} x \{competition models\} x \{bandit instances\}, and we are looking for findings that are consistent across this entire space. It is essential to keep each of the three dimensions small yet representative. In particular, we need to capture a huge variety of bandit algorithms and bandit instances with only a few representative examples. Further, we need a succinct and informative summarization of results within one atomic experiment and across multiple experiments (\eg see Table~\ref{sim_table}).
